Fast Sorting on Modern Computers

### 2005-08-27 docs/tech_report/elementary.tex
minor edits and merged in davids writing

\section{Bubblesort}
Bubblesort, as its name suggests, works by bubbling small keys from right of
the array to the left. It begins by taking a key from the end and moving it left
while it is less than any keys it passes. Once it comes across a smaller key, it
changes to this key, and moves it towards the left of the array. In this way, every
iteration the smallest key moves left to its final position, and other small
keys are slowly moved left, so that the unsorted part of the array becomes more
sorted as bubble sort continues.

@@

Finally, the branch prediction results are expected to be the same as in bubble
sort, with one difference. Keys which must travel the entire way across the
array are taken care of earlier in the sort. If the largest key is in the
leftmost position, it must be moved the entire way right, which causes misses
for every move in bubblesort, since each move is disjoint from the next. In
shakersort, however, each move in the correct direction will follow a move in the
same direction, each of which should be correctly predicted. This works for a
lot of keys, which are more likely to move in continuous movements than they are
in bubblesort, where only small keys on the right move continuously. The
performance should improve noticeably as a result.

@@

The branch behaviour of bubble sort is particularly interesting when
compared to selection sort. Although we usually think of the two
algorithms as being different, they are actually very similar. Like
selection sort, bubble sort finds the smallest (or largest) key in the
unsorted part of the array and moves it to start (end) of the array.
Thus, we would expect that bubble sort's branch prediction behaviour
would be similar to that of selection sort.

However, in the process of moving the smallest element into position,
bubble sort also rearranges other array elements. Selection sort
sweeps across the array and always keeps track of the left-to-right
minimum up to the current point. However, bubble sort actually moves
this minimum, exchanging pairwise with each element it encounters
which is larger. Thus, even the unsorted part of the array becomes
partially sorted. We might expect that bubble sort's comparison branch
for a partially sorted array might be more predictable than for a
random one, but it is not. With a random array, the smaller elements
are equally likely to appear anywhere in the array. With a partially
sorted array, the smaller elements tend to appear towards the start of
the array which is, in bubble sort, the last place that we look. Thus,
we change the candidate minimum much more often than if we were
searching a randomly ordered array.

>>>>

\section{Bubblesort}
Bubblesort, as its name suggests, works by bubbling small keys from right of
the array to the left. It begins by taking a key from the end and moving it left
while it is less than any keys it passes. Once it comes across a smaller key, it
changes to this key, and moves it towards the left of the array. In this way, every
iteration the smallest key moves left to its final position, and other small
keys are slowly moved left, so that the unsorted part of the array becomes more
sorted as bubblesort continues.

@@

Finally, the branch prediction results are expected to be the same as in
bubblesort, with one difference. Keys which must travel the entire way across
the array are taken care of earlier in the sort. If the largest key is in the
leftmost position, it must be moved the entire way right, which causes misses
for every move in bubblesort, since each move is disjoint from the next. In
shakersort, however, each move in the correct direction will follow a move in
the same direction, each of which should be correctly predicted. This works for
a lot of keys, which are more likely to move in continuous movements than they
are in bubblesort, where only small keys on the right move continuously. The
performance should improve noticeably as a result.

@@

The branch behaviour of bubblesort is particularly interesting when compared to
selection sort. Although we usually think of the two algorithms as being
different, they are actually very similar. Like selection sort, bubblesort finds
the smallest (or largest) key in the unsorted part of the array and moves it to
start (end) of the array.  Thus, we would expect that bubblesort's branch
prediction behaviour would be similar to that of selection sort.

However, in the process of moving the smallest element into position, bubblesort
also rearranges other array elements. Selection sort sweeps across the array and
always keeps track of the left-to-right minimum up to the current point.
However, bubblesort actually moves this minimum, exchanging pairwise with each
element it encounters which is larger. Thus, even the unsorted part of the array
becomes partially sorted. We might expect that bubblesort's comparison branch
for a partially sorted array might be more predictable than for a random one,
but it is not. With a random array, the smaller elements are equally likely to
appear anywhere in the array. With a partially sorted array, the smaller
elements tend to appear towards the start of the array which is, in bubblesort,
the last place that we look. Thus, we change the candidate minimum much more
often than if we were searching a randomly ordered array.

\begin{figure}
\includegraphics{plots/two_point_counter_bubblesort.eps}
\caption{Behaviour of branches in bubblesort compared to the ``sweep'' number}
\label{Predictability of branches in bubblesort compared to sweep number}
\end{figure}

To demonstrate this effect, we measured how the predictability of bubblesort's
comparison branch varies with successive sweeps across the array. We added a
simple software simulation of a bimodal predictor to the source code of base
bubblesort, and ran it on a large number of randomly generated arrays, each with
8192 elements.

In Figure \ref{Predictability of branches in bubblesort compared to sweep
number} we see that on the early iterations the number of correct predictions is
very close to the number of executed comparison branches. The comparison branch
is highly predictable because bubblesort's search for the smallest element is
similar to selection sort. However, with successive sweeps even the unsorted
part of the array becomes increasingly sorted and the branch becomes more
unpredictable. We see that the number of executed comparison branches falls
linearly, but that the number of correctly predicted comparison branches falls
much more quickly. Eventually, the array becomes so close to fully sorted that
the branch becomes increasingly predictable in the other direction, and the
number of correct predictions again approaches the number of executed branches.
However, during the intermediate stage, a huge number of branch mispredictions
have already occured. Paradoxically, by partially sorting the array, bubblesort
greatly increases the number of branch mispredictions and thus greatly increases
its running time. Bubblesort would be substantially faster if, instead of
attempting partially sort these intermediate elements, it simply left them
alone.

A particularly interesting result comes from a comparison between shakersort and
selection sort. The number of cycles to sort an array (shown in Figure
\ref{elementary cycles}) is nearly twice as high on shakersort as it is for
selection sort. Shakersort has a significantly lower instruction count and cache
miss count than selection sort, but has several orders of magnitude worse branch
misses. This shows how important it is to consider branch misses in an
algorithm, especially comparative ones. It also establishes that algorithms such
as bubblesort and shakersort, will perform considerably worse than algorithms
which have a correlation between successive branches, as in insertion and
selection sorts.
